Supporting information

### 2023-10-31 paper/main.qmd
Paper updates

---
title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism 
abstract: >
  This paper presents Maud, a command-line application that implements
  Bayesian statistical inference for kinetic models of biochemical metabolic
  reaction networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Below, we
  review the existing options in this area, explain how Maud improves on the state
  of the art, describe the intended modelling workflow and illustrate its use with
  an example application.
number-sections: true
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
    - id: biosustain
      name: DTU
      department: DTU Biosustain
      city: Kongens Lyngby
      country: Denmark
  - name: "Nicholas Luke Cowie"
    affiliations: 
    - ref: biosustain
  - name: "Lars Keld Nielsen"
    affiliations: 
    - ref: biosustain
    - id: UQ
    - name: The University of Queensland
    - department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
    - city: St Lucia
    - country: Australia
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
      abbreviations: IR,NMR,UV
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

The simulated dataset and underlying kinetic model that we used for our analysis
can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/methionine>
and is described in supplementary information section [REFERENCE].

@@

The prior distributions and corresponding true parameter values used in our
case study are shown in supplementary materials section [REFERENCE]. The first
two columns show the 1% and 99% quantiles of each marginal prior distribution.
True parameter value are shown in column three, and the last column shows the
z-score on log scale of the true parameter value according the marginal prior
distribution. As can be seen from the table, there are 7 parameters for which
the true value is outside the 1%-99% range. This is desirable, making the case
study more realistic, because extreme deviance from the prior distribution is
likely to occur in practice due to in vivo to in vitro measurement differences.

@@

We used the model of enzyme saturation from @liebermeister_modular_2010
and the generalised Monod-Wyman-Changeux model of Allosteric
regulation introduced in [@monod_nature_1965; @changeux_2013;
@popova_generalization_1975; @popova_description_1979] and used more recently
in @matosGRASPComputationalPlatform2022. To capture the effect on enzyme
activity of coupled phosphorylation and dephosphorylation processes we developed
a new mathematical model inspired by the generalised MWC model of allosteric
regulation. Full details of all mathematical aspects of Maud’s kinetic model can
be found in supplementary material section (REFERENCE).

@@

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See table (REFERENCE) in
this paper's supplementary materials for a description of all these parameters
and their dimensions. Note that some quantities in Maud’s kinetic model are
not treated as parameters: for example, temperatures, compartment volumes and
the formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values can be
used. Although in practice there can be considerable uncertainty regarding
these quantities, we chose to disregard this uncertainty in the interest of
simplicity.

@@

The document `supplementary.pdf` contains the following information:

- Description of Maud's kinetic model including parameter dimensions and rate
equations.
- Details of the procedure used to generate our case study results, including
generating the artificial case study datasets, specifying priors and carrying
out computation.
- Specification of the prior distribution used for the case studies

>>>>

---
title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism 
abstract: >
  This paper presents Maud, a command-line application that implements Bayesian
  statistical inference for kinetic models of biochemical metabolic reaction
  networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Our paper
  reviews the existing options in this area, presents a case study illustrating
  how Maud can be used to analyse a metabolic network and explains the
  biological, statistical and computational design decisions underpinning Maud.
number-sections: true
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
    - id: biosustain
      name: DTU
      department: DTU Biosustain
      city: Kongens Lyngby
      country: Denmark
  - name: "Nicholas Luke Cowie"
    affiliations: 
    - ref: biosustain
  - name: "Lars Keld Nielsen"
    affiliations: 
    - ref: biosustain
    - id: UQ
      name: The University of Queensland
      department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
      city: St Lucia
      country: Australia
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

The simulated dataset and underlying kinetic model that we used for our analysis
can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/methionine>
and is described in supporting information section 3.

@@

The prior distributions and corresponding true parameter values used in our
case study are shown in supporting information section 3.2. They were chosen to
reflect a plausible pre-experimental information state. In 7 cases, the marginal
prior distribution for a parameter disagrees with the true parameter values used
to generate the data. A similar situation is likely to occur in practice due to
*in vivo* vs *in vitro* measurement differences.

@@

We used the model of enzyme saturation from @liebermeister_modular_2010
and the generalised Monod-Wyman-Changeux model of Allosteric
regulation introduced in [@monod_nature_1965; @changeux_2013;
@popova_generalization_1975; @popova_description_1979] and used more recently
in @matosGRASPComputationalPlatform2022. To capture the effect on enzyme
activity of coupled phosphorylation and dephosphorylation processes we developed
a new mathematical model inspired by the generalised MWC model of allosteric
regulation. Full details of all mathematical aspects of Maud’s kinetic model can
be found in supporting information section 2.

@@

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See table (REFERENCE) in
this paper's supporting information for a description of all these parameters
and their dimensions. Note that some quantities in Maud’s kinetic model are
not treated as parameters: for example, temperatures, compartment volumes and
the formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values can be
used. Although in practice there can be considerable uncertainty regarding
these quantities, we chose to disregard this uncertainty in the interest of
simplicity.

@@

The analyses described in this paper, as well as instructions for reproducing
our results, figures and paper, can be found at <https://biosustain/Methionine_model>.

The document `supporting_information.pdf` contains the following information:

- Description of Maud's input format
- Description of Maud's kinetic model including rate equations as well as
parameters and their dimensions.
- Details of the procedure used to generate our case study results, including
generating the artificial case study datasets, specifying priors and carrying
out computation.
- Specification of the prior distribution used for the case studies.

### 2023-10-31 paper/main.tex
Paper updates

\author{Lars Keld Nielsen}
\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
\affiliation{   }

@@

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

@@

The simulated dataset and underlying kinetic model that we used for our
analysis can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/data/methionine}
and is described in supplementary information section {[}REFERENCE{]}.

@@

The prior distributions and corresponding true parameter values used in
our case study are shown in supplementary materials section
{[}REFERENCE{]}. The first two columns show the 1\% and 99\% quantiles
of each marginal prior distribution. True parameter value are shown in
column three, and the last column shows the z-score on log scale of the
true parameter value according the marginal prior distribution. As can
be seen from the table, there are 7 parameters for which the true value
is outside the 1\%-99\% range. This is desirable, making the case study
more realistic, because extreme deviance from the prior distribution is
likely to occur in practice due to in vivo to in vitro measurement
differences.

@@

MCMC sampling for our comparison input yielded 200 samples in 625
minutes; Laplace sampling yielded the same number of samples in only one
minute. The diagnostics indicated that our algorithm was able to find
the maximum a posteriori parameter configuration, approximate the
Hessian and use these quantities to generate approximate posterior
samples. The results can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

@@

\caption{\label{fig-missing}Results of removing concentration
measurements for the metabolite \(ahcys_c\) from our case study dataset.
(\textbf{Top}) A comparison of metabolite concentration residuals
between the full measurement dataset (blue-green) and the missing-data
dataset (grey), displayed on natural logarithmic scale. The missing-data
model was unable to predict the withheld \(ahcys_c\) concentrations.
(\textbf{Bottom Left}) The marginal posterior distribution for the
Michaelis constant \(K_m^{AHC1,ahcys_c}\) in each model, alongside the
true parameter value used to generate both datasets. The true value is
recovered by the complete-data model but not by the missing-data model.
(\textbf{Bottom Right}) The distribution of total log-likelihood for
out-of-sample flux measurements in both models. There is a significant
overlap between the two distributions, suggesting that removing the
\(ahcys_c\) measurement did not cause catastrophic prediction failure.
However, overall the complete-data model tended to make better
predictions.}

@@

Maud is a command line application implementing Bayesian inference for a
wide range of realistic kinetic models. Maud is written in Python
\citep{vanrossumPythonReferenceManual2009}, designed for use on Windows,
macOS and Linux, registered on the Python Package Index as
\texttt{maud-metabolic-models}, documented at
\url{https://maud-metabolic-models.readthedocs.io} and actively
developed and maintained at \url{https://github.com/biosustain/Maud/}.

To use Maud, a user must first collate appropriate input information,
represent it in files with Maud's required formats (see section
(REFERENCE)below). Maud's command line interface provides commands for
inference, simulation and making out-of-sample predictions. Results are
stored in files, using a structured, interoperable format.

\hypertarget{input-format}{%
\subsection{Input format}\label{input-format}}

Maud inputs are structured directories, somewhat inspired by the PEtab
format \citep{SchmiesterSch2021}. A Maud input directory must contain a
toml \citep{preston-wernertomandgedampradyunTOMLSpecification0rc2020}
file called \texttt{config.toml} which gives the input a name,
configures how Maud will be run and tells Maud where to find the other
files, allowing these to have custom names. It must also include a file
containing a kinetic model definition, a file specifying information
about parameters and a file with information experiments. The required
structure of these files is documented at \textless https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html\textgreater.
The input is validated against a Pydantic
\citep{pydanticdevelopersPydantic2022} data model which can be inspected
at \url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.

We chose to implement a custom input format despite the existence of
standard formats in similar areas, including SBML
\citep{keatingSBMLLevelExtensible2020} and PEtab
\citep{SchmiesterSch2021}. This choice was partly motivated by the need
to ensure flexibility as Maud was developed, but there are also features
of SBML and PEtab that make them structurally unsuitable in this
context. Our requirements for an input format included that it be
mathematics-free, so that all mathematical details are encapsulated in
source code, and that it has a detailed, verifiable structure. These
requirements made toml more attractive than SBML: toml is easier for
humans to read and edit and can straightforwardly be validated using
tools like Pydantic. Further, an SBML representation of our desired
input would not contain differential equations. It would therefore not
be interoperable with most SBML targeting software, which typically
assumes that differential equations are available and does not know
about Maud's structure.

@@

We used the model of enzyme saturation from
\citet{liebermeister_modular_2010} and the generalised
Monod-Wyman-Changeux model of Allosteric regulation introduced in
\citep{monod_nature_1965, changeux_2013, popova_generalization_1975, popova_description_1979}
and used more recently in \citet{matosGRASPComputationalPlatform2022}.
To capture the effect on enzyme activity of coupled phosphorylation and
dephosphorylation processes we developed a new mathematical model
inspired by the generalised MWC model of allosteric regulation. Full
details of all mathematical aspects of Maud's kinetic model can be found
in supplementary material section (REFERENCE).

@@

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See table (REFERENCE) in this paper's supplementary materials
for a description of all these parameters and their dimensions. Note
that some quantities in Maud's kinetic model are not treated as
parameters: for example, temperatures, compartment volumes and the
formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values
can be used. Although in practice there can be considerable uncertainty
regarding these quantities, we chose to disregard this uncertainty in
the interest of simplicity.

@@

Maud uses the Python library click
\citep{clickdevelopersClickPythonComposable2022} to implement a command
line interface. The command line interface loads input files as Python
dictionaries, which are parsed using the Python library \texttt{toml}
\citep{pearsonTomlPythonLibrary2020} and then validated and converted
into structured \texttt{MaudInput} objects using Pydantic
\citep{pydanticdevelopersPydantic2022}. Maud's statistical model is
implemented in the probabilistic programming language Stan
\citep{carpenterStanProbabilisticProgramming2017} and accessed using the
interface cmdstanpy \citep{standevelopmentteamCmdStanPy2022}. For
posterior sampling, Maud uses the \texttt{MaudInput} to create a Stan
input json file and obtain configuration information for cmdstanpy,
which it uses to (if necessary) generate a model executable file and
then trigger posterior sampling using adaptive Hamiltonian Monte Carlo.
When sampling is complete, Maud converts to the output into the standard
format \texttt{InferenceData} using the Python library arviz
\citep{kumarArviZUnifiedLibrary2019} and saves it as a json file, along
with some information for debugging.

Two details of Maud's implementation are important to highlight: the
method of posterior sampling and how Maud solves the steady state
problem \eqref{eq-3}.

\hypertarget{posterior-sampling}{%
\subsubsection{Posterior sampling}\label{posterior-sampling}}

Although integrals of the joint probability model for kinetic models are
typically analytically intractable, they can be approximated numerically
using Markov Chain Monte Carlo (MCMC) and other methods. Maud uses MCMC
primarily because there exist many methods for verifying that MCMC
samples really do approximate the target probability distribution: see
\citet{vehtariRankNormalizationFoldingLocalization2021} and
\citet{taltsValidatingBayesianInference2018} for discussion of this
point. In addition, there are several examples of successful Bayesian
kinetic modelling projects using MCMC including
\citet{st.johnBayesianInferenceMetabolic2018} and
\citet{xingModelingKineticsLargescale2010}.

\hypertarget{solving-the-steady-state-problem}{%
\subsubsection{Solving the steady state
problem}\label{solving-the-steady-state-problem}}

In order to implement adaptive Hamiltonian Monte Carlo, Maud must
repeatedly solve the steady state problem \eqref{eq-3} and find its
gradients with respect to all parameters. To achieve this, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
\citep{serbanCVODESSensitivityEnabledODE2005}: CVODES and IDAS. These
solvers are accessed via their interface from Stan. The hybrid method
follows that proposed by \citet{margossianComputingSteadyStates2018},
and involves numerically evolving the ODE system for a short period of
time, then using the difference between the evolved and starting
concentrations as the target for a numerical algebra solver.

@@

The document \texttt{supplementary.pdf} contains the following
information:

\begin{itemize}
\tightlist
\item
  Description of Maud's kinetic model including parameter dimensions and
  rate equations.
\item
  Details of the procedure used to generate our case study results,
  including generating the artificial case study datasets, specifying
  priors and carrying out computation.
\item
  Specification of the prior distribution used for the case studies
\end{itemize}

>>>>

\author{Lars Keld Nielsen}
\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
\affiliation{Australian Institute for Bioengineering and Nanotechnology
(AIBN), The University of Queensland, St Lucia, Australia}

@@

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

@@

The simulated dataset and underlying kinetic model that we used for our
analysis can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/data/methionine}
and is described in supporting information section 3.

@@

The prior distributions and corresponding true parameter values used in
our case study are shown in supporting information section 3.2. They
were chosen to reflect a plausible pre-experimental information state.
In 7 cases, the marginal prior distribution for a parameter disagrees
with the true parameter values used to generate the data. A similar
situation is likely to occur in practice due to \emph{in vivo} vs
\emph{in vitro} measurement differences.

@@

MCMC sampling for our comparison input yielded 800 samples in 625
minutes; Laplace sampling yielded the same number of samples in only one
minute. The diagnostics indicated that our algorithm was able to find
the maximum a posteriori parameter configuration, approximate the
Hessian and use these quantities to generate approximate posterior
samples. The results can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

@@

\caption{\label{fig-missing}Results of removing concentration
measurements for the metabolite \(ahcys_c\) from our case study dataset.
(\textbf{Top}) A comparison of metabolite concentration residuals
between the full measurement dataset (blue-green) and the missing-data
dataset (grey), displayed on natural logarithmic scale. The missing-data
model was unable to estimate the withheld \(ahcys_c\) concentrations.
(\textbf{Bottom Left}) The marginal posterior distribution for the
Michaelis constant \(K_m^{AHC1,ahcys_c}\) in each model, alongside the
true parameter value used to generate both datasets. The true value is
recovered by the complete-data model but not by the missing-data model.
(\textbf{Bottom Right}) The distribution of total log-likelihood for
out-of-sample flux measurements in both models. There is a significant
overlap between the two distributions, suggesting that removing the
\(ahcys_c\) measurement did not cause catastrophic prediction failure.
However, overall the complete-data model tended to make better
predictions.}

@@

Maud is a command line application implementing Bayesian inference for a
wide range of realistic kinetic models. Maud is written in Python
\citep{vanrossumPythonReferenceManual2009}, designed for use on Windows,
macOS and Linux, is \texttt{pip} installable from the Python Package
Index as \texttt{maud-metabolic-models}, documented at
\url{https://maud-metabolic-models.readthedocs.io} and actively
developed and maintained at \url{https://github.com/biosustain/Maud/}.

To use Maud, a user must first collate appropriate input information,
represent it in files with Maud's required formats (see supporting
information section on input format). Maud's command line interface
provides commands for inference via a range of algorithms including
adaptive Hamiltonian Monte Carlo, as well as commands for simulation and
making out-of-sample predictions. Results are stored in files, using a
structured, interoperable format.

As well as parameter values, Maud also performs inference for derived
quantities including the components of the regulatory decomposition
described below in equation \eqref{eq-decomposition}, log likelihoods,
simulated measurement values and metabolic control analysis coefficients
\citep{kacser_control_1973}.

In the rest of this section we describe Maud's kinetic model at high
level and discuss Maud's statistical model, implementation and how it
solves the crucial steady state problem. Further details about Maud's
kinetic model can be found in the supporting information.

@@

We used the model of enzyme saturation from
\citet{liebermeister_modular_2010} and the generalised
Monod-Wyman-Changeux model of Allosteric regulation introduced in
\citep{monod_nature_1965, changeux_2013, popova_generalization_1975, popova_description_1979}
and used more recently in \citet{matosGRASPComputationalPlatform2022}.
To capture the effect on enzyme activity of coupled phosphorylation and
dephosphorylation processes we developed a new mathematical model
inspired by the generalised MWC model of allosteric regulation. Full
details of all mathematical aspects of Maud's kinetic model can be found
in supporting information section 2.

@@

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See table (REFERENCE) in this paper's supporting information
for a description of all these parameters and their dimensions. Note
that some quantities in Maud's kinetic model are not treated as
parameters: for example, temperatures, compartment volumes and the
formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values
can be used. Although in practice there can be considerable uncertainty
regarding these quantities, we chose to disregard this uncertainty in
the interest of simplicity.

@@

Maud uses the Python library click
\citep{clickdevelopersClickPythonComposable2022} to implement a command
line interface. The command line interface loads input files as Python
dictionaries, which are parsed using the Python library \texttt{toml}
\citep{pearsonTomlPythonLibrary2020} and then validated and converted
into structured \texttt{MaudInput} objects using Pydantic
\citep{pydanticdevelopersPydantic2022}. Maud's statistical model is
implemented in the probabilistic programming language Stan
\citep{carpenterStanProbabilisticProgramming2017} and accessed using the
interface cmdstanpy \citep{standevelopmentteamCmdStanPy2022}. For
posterior sampling, Maud uses the \texttt{MaudInput} to create an input
file for Stan and then trigger posterior sampling using adaptive
Hamiltonian Monte Carlo. See
\citet{betancourtConceptualIntroductionHamiltonian2018} for more about
this algorithm.

When sampling is complete, Maud converts to the output into the standard
format \texttt{InferenceData} using the Python library arviz
\citep{kumarArviZUnifiedLibrary2019} and saves it as a json file, along
with some information for debugging. This format allows for easy
checking of MCMC diagnostics including divergent transitions, effective
sample size and the improved \(\hat{R}\) statistic proposed in
\citet{vehtariRankNormalizationFoldingLocalization2021}.

\hypertarget{solving-the-steady-state-problem}{%
\subsection{Solving the steady state
problem}\label{solving-the-steady-state-problem}}

Maud's parameters are connected with measurable quantities via the
steady state equation \eqref{eq-steady}. Posterior sampling using
adaptive Hamiltonian Monte Carlo requires repeatedly solving equation
\eqref{eq-steady} and finding the gradients of its solution with respect
to all model parameters. This must be done numerically as analytic
solutions are not available for realistic kinetic models.

The speed at which this problem can be solved is tightly coupled with
the size and complexity of metabolic network that can practically be
modelled. See \citet{timonenImportanceSamplingApproach2022a} for more
about considerations involved in this kind of modelling.

To solve equation \eqref{eq-steady} and find its gradients, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
\citep{serbanCVODESSensitivityEnabledODE2005}: CVODES and IDAS via their
interface from Stan. The hybrid method follows that proposed by
\citet{margossianComputingSteadyStates2018}, and involves numerically
evolving the ODE system for a short period of time, then using the
difference between the evolved and starting concentrations as the target
for a numerical algebra solver.

@@

The analyses described in this paper, as well as instructions for
reproducing our results, figures and paper, can be found at
\url{https://biosustain/Methionine_model}.

The document \texttt{supporting\_information.pdf} contains the following
information:

\begin{itemize}
\tightlist
\item
  Description of Maud's input format
\item
  Description of Maud's kinetic model including rate equations as well
  as parameters and their dimensions.
\item
  Details of the procedure used to generate our case study results,
  including generating the artificial case study datasets, specifying
  priors and carrying out computation.
\item
  Specification of the prior distribution used for the case studies.
\end{itemize}

### 2023-10-31 paper/supporting_information.qmd
Paper updates

---
title: Supporting information
bibliography: bibliography.bib
format:
  html:
    embed-resources: true
    include-in-header: mathjax.html
  pdf: default
  docx: default
---

@@

Maud inputs are structured directories, somewhat inspired by the PEtab
format [@SchmiesterSch2021]. A Maud input directory must contain a toml
[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
`config.toml` which gives the input a name, configures how Maud will be run
and tells Maud where to find the other files, allowing these to have custom
names. It must also include a file containing a kinetic model definition,
a file specifying information about parameters and a file with information
experiments. The required structure of these files is documented at <https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.

@@

@tbl-case-study-params shows the prior distributions we used for independent
parameters.

@@

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
-data datasets. For the full dataset we obtained 1000 post-warmup samples each
from 4 independent Markov chains after 1000 warm-up samples and "hot-starting"
with a mass metric output by a previous model run.

@@

To compare MCMC sampling with Laplace approximation we used a different model.
 The full Maud input folders can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode> 
and <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode_laplace>

>>>>

---
title: Supporting information
bibliography: bibliography.bib
number-sections: true
format:
  html:
    embed-resources: true
    include-in-header: mathjax.html
  pdf: default
  docx: default
---

This document provides information in support of our article "Bayesian
regression facilitates quantitative modelling of cell metabolism".

@@

Maud inputs are structured directories, somewhat inspired by the PEtab
format [@SchmiesterSch2021]. A Maud input directory must contain a toml
[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
`config.toml` which gives the input a name, configures how Maud will be run
and tells Maud where to find the other files, allowing these to have custom
names. It must also include a file containing a kinetic model definition,
a file specifying information about parameters and a file with information
experiments. The required structure of these files is documented at <https://maud-metabolic-models.readthedocs.io/en/latest/inputting.html>.
The input is validated against a Pydantic [@pydanticdevelopersPydantic2022] 
data model.

@@

@tbl-case-study-params shows the prior distributions we used for independent
parameters. The first two columns show the 1% and 99% quantiles of each
marginal prior distribution. True parameter value are shown in column three,
and the last column shows the z-score on log scale of the true parameter value
according the marginal prior distribution. As can be seen from the table, there
are 7 parameters for which the true value is outside the 1%-99% range.

@@

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
missing-data datasets. For the full dataset we obtained 1000 post-warmup
samples each from 4 independent Markov chains after 1000 warm-up samples and
“hot-starting” with a mass metric output by a previous model run.

@@

To compare MCMC sampling with Laplace approximation we used a different model
with fewer parameters and state variables. This model was chosen because
we were not able to generate results for our methionine model using Laplace
approximation. The simpler case still serves to illustrate the general issues
with approximating the posterior distributions of Bayesian kinetic models using
the Laplace method, and that the associated numerical instability is another
reason to prefer other methods where possible.

The full Maud input folders used for our Laplace approximation case study can
be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode>
and <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode_laplace>.
